---
title: "Learning Data Analysis"
format: 
  html:
    embed-resources: true
---

```{r}
#| message: false
library(tidyverse)
library(ggplot2)
```

# Reading data file

## Text-based data (csv)

:::{.callout-note}

This is how to read CSV file into R.

:::

```{r}
#| message: false
#| warning: false
data <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2025/2025-04-15/penguins.csv')
data
```

:::{.callout-note}

This is how to write a CSV file in R.

:::

```{r}
write_csv(data, here::here("data/penguins-r.csv"))
```


# Data Manipulation

## Manipulating Variables

:::{.callout-note}

This is how to select variables in R.

:::

```{r}
data |>
  select(species, island, sex)
```

:::{.callout-note}

This is how to rename variables in R.

:::

```{r}
data |>
  rename(body_mass_kg = body_mass)
```

:::{.callout-note}

This is how to compute variables in R.

:::

```{r}
data |>
  mutate(body_mass_kg = body_mass/1000)
```


## Manipulating Observations

:::{.callout-note}

This is how you select rows in R.

:::

```{r}
data |> 
  filter(species == "Adelie")
```

:::{.callout-note}

This is how to get unique rows in R.

:::

```{r}
data |> 
  distinct(species, island)
```

:::{.callout-note}

This is how to arrange the row in R.

:::

```{r}
data |> 
  arrange(desc(bill_len))
```


## Grouped operation

:::{.callout-note}
This is how to group data by one or more variables in R.
:::

```{r}
data |> 
  group_by(species) |> 
  mutate(above_average_mass = body_mass > mean(body_mass, na.rm = TRUE)) |> 
  ungroup()
```

## Summarising Observations

```{r}
data |> 
  summarise(
    min_flipper = min(flipper_len, na.rm = TRUE),
    median_flipper = median(flipper_len, na.rm = TRUE),
    max_flipper = max(flipper_len, na.rm = TRUE)
  )

data |>
  group_by(species) |>
  summarise(average_bill_length = mean(bill_len, na.rm = TRUE))

data |> 
  count(species, island)
```


## Joining Tables

:::{.callout-note}

This is how you combine data in R.

:::

```{r}
#| code-fold: true
split <- data |> 
  group_by(year) |> 
  group_split()

data_07 <- split[[1]]
data_08 <- split[[2]]
```

```{r}
bind_rows(data_07, data_08)
```

:::{.callout-note}

This is how to join the data in R.

:::

```{r}
penguin_info <- tibble(
  species = c("Adelie", "Chinstrap", "Gentoo"),
  scientific_name = c("Pygoscelis adeliae", "Pygoscelis antarcticus", "Pygoscelis papua"),
  nest_type = c("Nests made from stones", "Nests made from stones", "Nests lined with pebbles and vegetation"),
  conservation_status = c("Least Concern", "Least Concern", "Least Concern")
)
penguin_info
```

```{r}
data |>
  left_join(penguin_info, by = "species")
```


# Data Visualisation

In R, we use ggplot2.

```{r}
library(ggplot2)

data |> 
  ggplot(aes(x = bill_len, y = body_mass, color = species)) +
  geom_point()
```






